SunPy - Python for Solar Physics

### 2014-01-28 2-1-Map.tex

\subsection{Map}\label{ssec:map}
The Map data type represents 2D spatial data such as images of the Sun and 
inner heliosphere. It provides a wrapper around a data array and its associated
spatial coordinates and other metadata. The \texttt{Map} class provides 
methods for typical operations on 2D data, 
such as rotation and re-sampling, and visualisation functionality.
The \texttt{Map} class
also provides a convenient interface for loading data from a variety of sources, typically
a FITS file as shown in Listing~\ref{code:aia_1}.

The architecture of the \texttt{Map} class consists of a template map called
\texttt{GenericMap}, which is a subclass of \texttt{astropy.nddata.NDData}. \texttt{NDData} is a 
generic wrapper around an \texttt{ndarray} with a \texttt{.meta} attribute to store metadata.
As \texttt{NDData} is currently still in development, \texttt{GenericMap} does not yet make full
use of its capabilities, but this inheritance structure provides 
for future integration of \texttt{Astropy}. In order to provide instrument- or
detector-specific integration, \texttt{GenericMap} is subclassed on creation. 
Each subclass of \texttt{GenericMap} can register with the \texttt{Map} creation function (\textit{i.e.}, factory)
by implementing a method that returns \texttt{True} if the metadata
in the file header matches
that instrument or detector.  This \texttt{Map} factory 
will then automatically return an instance of the specific \texttt{GenericMap} 
subclass. SunPy 0.4 has \texttt{Map} specialisations for the 
following instruments: 
\textit{YOHKOH/SXT}, \textit{SOHO/EIT} and \textit{LASCO}, \textit{RHESSI}, 
\textit{STEREO/EUVI} and \textit{COR}, \textit{HINODE/XRT},
\textit{PROBA2/SWAP}, \textit{SDO/AIA} and \textit{HMI}, 
and \textit{IRIS} SJI frames.

@@

In addition to the data-type classes, the \texttt{map} sub-package provides two 
collection classes, \texttt{CompositeMap} and \texttt{MapCube}, for 
temporally and spatially aligned data respectively.
\texttt{CompositeMap} provides methods for overlaying spatially aligned 
data, with support for visualisation of images and contour lines overlaid 
upon each other.
\texttt{MapCube} 
provides methods for animation of its series of \texttt{Map} objects. 
Listings~\ref{code:compmap_1} and \ref{code:mapcube_1} show how to interact 
with these classes.
In addition, Listing~\ref{code:mapcube_2} shows how the \texttt{MapCube} 
classes interacts with the \texttt{animation} module to save a video.

>>>>

\subsection{Map}\label{ssec:map}
The Map data type represents 2D spatial data, such as images of the Sun and 
inner heliosphere. It provides a wrapper around a \texttt{numpy} data array and 
its associated spatial coordinates and other metadata. The \texttt{Map} class 
provides methods for typical operations on 2D data, such as rotation, 
re-sampling, and visualisation functionality.
The \texttt{Map} class also provides a convenient interface for loading data 
from a variety of sources, typically a FITS file as shown in 
Listing~\ref{code:aia_1}.

The architecture of the \texttt{Map} class consists of a template map called
\texttt{GenericMap}, which is a subclass of \texttt{astropy.nddata.NDData}. 
\texttt{NDData} is a generic wrapper around an \texttt{ndarray} with a 
\texttt{.meta} attribute to store metadata.
As \texttt{NDData} is currently still in development, \texttt{GenericMap} does 
not yet make full use of its capabilities, but this inheritance structure 
provides for future integration of \texttt{Astropy}. In order to provide 
instrument- or detector-specific integration, \texttt{GenericMap} is designed 
in such a way that it can be subclassed to implement instrument specific 
functionality. Each of these subclasses of \texttt{GenericMap} can register 
with the \texttt{Map} creation factory by implementing a method that returns 
\texttt{True} if the metadata in the file header matches that instrument or 
detector.  This \texttt{Map} factory will then automatically return an instance 
of the specific \texttt{GenericMap} subclass dependant upon the data provided. 
SunPy 0.4 has \texttt{Map} specialisations for the following instruments: 
\textit{YOHKOH/SXT}, \textit{SOHO/EIT} and \textit{LASCO}, \textit{RHESSI}, 
\textit{STEREO/EUVI} and \textit{COR}, \textit{HINODE/XRT},
\textit{PROBA2/SWAP}, \textit{SDO/AIA} and \textit{HMI}, 
and \textit{IRIS} SJI frames.

@@

In addition to the data-type classes, the \texttt{map} sub-package provides two 
collection classes, \texttt{CompositeMap} and \texttt{MapCube}, for 
temporally and spatially aligned data respectively.
\texttt{CompositeMap} provides methods for overlaying spatially aligned 
data, with support for visualisation of images and contour lines overlaid 
upon each other.
\texttt{MapCube} 
provides methods for animation of its series of \texttt{Map} objects. 
Listings~\ref{code:compmap_1} and \ref{code:mapcube_1} show how to interact 
with these classes.
In addition, Listing~\ref{code:mapcube_2} shows how the \texttt{MapCube} 
classes interacts with the \texttt{matplotlib.animation} submodule to save a 
video.

### 2014-01-28 2-2-Lightcurve.tex

Time series data and their analysis are a fundamental aspect of solar
physics for which many data sources are available.
SunPy provides a \texttt{LightCurve} class
with a convenient and consistent interface for handling solar time-series
data.  The main engine behind the \texttt{LightCurve} class is
the \texttt{pandas} data-analysis library (\url{http://pandas.pydata.org}),
and \texttt{LightCurve}'s
\texttt{.data} attribute is a \texttt{pandas.DataFrame} class.
The \texttt{pandas} library contains a large amount
of functionality for manipulating and analysing time-series data,
making it an ideal basis for \texttt{LightCurve}.  \texttt{LightCurve}
assumes that the input data are time-ordered list(s) of numbers, and each
list becomes a column in the \texttt{pandas} DataFrame class.

Currently, the \texttt{LightCurve} class is compatible with the
following data sources: the GOES X-ray Sensor (XRS), PROBA2/LYRA, and
the SDO EUV Variability Experiment (EVE; only the level ``OCS'' and
% Do we need to explain what OCS means?
average CSV files -- see \url{http://lasp.colorado.edu/home/eve/data/}
for more detail).  For each of these instruments, a subclass of the
\texttt{LightCurve} object is initialised
(\textit{e.g.}, \texttt{GOESLightCurve}) which inherits from
\texttt{LightCurve} but allows instrument-specific functionality to be
included.  Future developments will introduce support for additional
instruments and data products.  Since there is no established standard
as to how time-series data
should be stored and distributed, each SunPy \texttt{LightCurve}
object provides the ability to download its corresponding specific
data format in its constructor.

%\subsubsection{Creating and using Lightcurves}

A \texttt{LightCurve} object may be created using a number of different methods. 
Firstly, a \texttt{LightCurve} may be created for a specific instrument given
an input time range. In Listing~\ref{code:goes_lc}, 
the \texttt{LightCurve} constructor searches a remote site for the GOES X-ray data 
specified by the time interval, downloads the required files and subsequently 
creates the object.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import lightcurve
>>> from sunpy.time import TimeRange
>>> goes = lightcurve.GOESLightCurve.create('2011-06-07 06:00',
...                                         '2011-06-07 10:00')
>>> goes.peek()
>>> trange = TimeRange('2011-06-07 06:00','2011-06-07 07:00')
%>>> goes_zoom = goes.truncate(trange)
% the above does not work!
>>> resampled_goes = goes.data.resample("1H")
% Does 1h resample makes sense?? I vote for an example with some utility
\end{minted}
\begin{center}
\includegraphics[width=10cm]{goes_lightcurve.pdf}
\end{center}
\caption{Example retrieval of a GOES lightcurve for the time interval 06:00--08:00 UT on 
2011 June 7 using a time range, and the output of the \texttt{peek()} method.
A sub-interval is then extracted from the Lightcurve and finally the data is resampled
to 1 hour cadence.}
\label{code:goes_lc}
\end{listing}
%schriste - note to self, fix this example, add resampled points on top of plot?

@@

>>>>

Time series data and their analysis are a fundamental aspect of solar
physics for which many data sources are available.
SunPy provides a \texttt{LightCurve} class
with a convenient and consistent interface for handling solar time-series
data.  The main engine behind the \texttt{LightCurve} class is
the \href{http://pandas.pydata.org}{\texttt{pandas}} data-analysis library, and 
\texttt{LightCurve}'s \texttt{.data} attribute is a \texttt{pandas.DataFrame} 
class.
The \texttt{pandas} library contains a large amount
of functionality for manipulating and analysing time-series data,
making it an ideal basis for \texttt{LightCurve}.  \texttt{LightCurve}
assumes that the input data are time-ordered list(s) of numbers, and each
list becomes a column in the \texttt{pandas} DataFrame class.

Currently, the \texttt{LightCurve} class is compatible with the
following data sources: the GOES X-ray Sensor (XRS), PROBA2/LYRA, and
the SDO EUV Variability Experiment (EVE; only the level ``OCS'' and
% Do we need to explain what OCS means?
average CSV files -- see \url{http://lasp.colorado.edu/home/eve/data/}
for more detail).  For each of these instruments, a subclass of the
\texttt{LightCurve} object is initialised
(\textit{e.g.}, \texttt{GOESLightCurve}) which inherits from
\texttt{LightCurve} but allows instrument-specific functionality to be
included.  Future developments will introduce support for additional
instruments and data products, as well as implementing a factory interface 
similar to that of \texttt{Map}.  Since there is no established standard
as to how time-series data should be stored and distributed, each SunPy 
\texttt{LightCurve} object provides the ability to download its corresponding 
specific data format in its constructor and parse that file type.

A \texttt{LightCurve} object may be created using a number of different methods. 
Firstly, a \texttt{LightCurve} may be created for a specific instrument given
an input time range. In Listing~\ref{code:goes_lc}, 
the \texttt{LightCurve} constructor searches a remote source for the GOES X-ray 
data specified by the time interval, downloads the required files and 
subsequently creates and plots the object.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy import lightcurve
>>> from sunpy.time import TimeRange
>>> goes = lightcurve.GOESLightCurve.create('2011-06-07 06:00',
...                                         '2011-06-07 10:00')
>>> goes.peek()
>>> resampled_goes = goes.data.resample("1H")
% Does 1h resample makes sense?? I vote for an example with some utility
% an example is an example right.
\end{minted}
\begin{center}
\includegraphics[width=10cm]{goes_lightcurve.pdf}
\end{center}
\caption{Example retrieval of a GOES lightcurve for the time interval 
06:00--08:00 UT on 2011 June 7 using a time range, and the output of the 
\texttt{peek()} method.
Finally the data is resampled to a 1 hour cadence.}
\label{code:goes_lc}
\end{listing}
%schriste - note to self, fix this example, add resampled points on top of plot?

@@

%Stuart: Where did the pandas section go?

### 2014-01-28 2-3-Spectra.tex

The \texttt{Spectrogram} class supports radio spectrograms from the e-Callisto 
solar radio spectrometer network (\url{http://www.e-callisto.org/})
and STEREO/SWAVES spectrograms.
As with other SunPy data types, the \texttt{Spectrogram} class has been
built so that each instrument initialises using a subclass containing the instrument-specific 
functionalities.
%The common functionality provided by the base \texttt{Spectrogram} class includes
%reading data,
%joining different time ranges and frequencies,
%performing frequency-dependent background subtraction,
%and convenient visualization and sampling of the data.

@@

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.spectra.sources.callisto import CallistoSpectrogram
>>> tstart, tend = "2011-06-07T06:00:00", "2011-06-07T07:45:00"
>>> callisto = CallistoSpectrogram.from_range("BIR", tstart, tend)
>>> callisto_nobg = callisto.subtract_bg()
>>> callisto_nobg.peek(vmin = 0)
\end{minted}
\begin{center}
\includegraphics[width=0.8\columnwidth]{callisto_nobg}
\end{center}
\caption{Example of how \texttt{CallistoSpectrogram} retrieves the
  data for the requested time range and observatory, merges it and
  removes the background signal.  The data requested -- 'BIR' -- is
  the code name of the Rosse Observatory at Birr Castle in Ireland
  (\url{http://www.rosseobservatory.ie})}
\label{code:spectra}
\end{listing}

>>>>

The \texttt{Spectrogram} class supports radio spectrograms from the 
\href{http://www.e-callisto.org/}{e-Callisto}
solar radio spectrometer network and STEREO/SWAVES spectrograms.
As with other SunPy data types, the \texttt{Spectrogram} class has been
built so that each instrument initialises using a subclass containing the instrument-specific 
functionalities.
%The common functionality provided by the base \texttt{Spectrogram} class includes
%reading data,
%joining different time ranges and frequencies,
%performing frequency-dependent background subtraction,
%and convenient visualization and sampling of the data.

@@

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.spectra.sources.callisto import CallistoSpectrogram
>>> tstart, tend = "2011-06-07T06:00:00", "2011-06-07T07:45:00"
>>> callisto = CallistoSpectrogram.from_range("BIR", tstart, tend)
>>> callisto_nobg = callisto.subtract_bg()
>>> callisto_nobg.peek(vmin = 0)
\end{minted}
\begin{center}
\includegraphics[width=0.8\columnwidth]{callisto_nobg}
\end{center}
\caption{Example of how \texttt{CallistoSpectrogram} retrieves the
  data for the requested time range and observatory, merges it and
  removes the background signal.  The data requested -- `BIR' -- is
  the code name of the \href{http://www.rosseobservatory.ie}{Rosse Observatory}
  at Birr Castle in Ireland.}
\label{code:spectra}
\end{listing}

### 2014-01-28 3-5-Database.tex

Easy access to large quantities of solar data frequently leads to datafiles accumulating
in local storage such as laptops and desktop computers. Keeping data organised and available
is typically a cumbersome task for the average user. The file database is a sub-package of 
SunPy which is meant to solve this problem by providing a unified database to store and manage information about 
local data files. The database package was developed as part of GSOC 2013.

The database sub-package makes use of a database supported by
SQLAlchemy (\url{http://www.sqlalchemy.org}). This library was chosen
since it supports many SQL dialects. 
If SQLite is chosen, the database is stored as a single file which is
created automatically. This solution makes it possible to easily backup, copy and distribute
a database file. A server-based database, on the other hand, could be used by a collaboration 
who work together from different machines on the
same data: a central database server stores all data and the clients connect to
it to read or write data.

The database can store and manage all data that is retrieved via SunPy from the net
package though currently only the \textsc{VSO} package is supported. 
It also possible to add entries manually by file or directory. The package also provides
a \texttt{fetch} method which provides a unified data search which includes both local files
and files on the \textsc{VSO} are searched. This reduces the likelihood of downloading the same file 
multiple times. When a file is added to the database the file is scanned and a file hash is produced. 
The current date is associated with the entry along with metadata summaries such 
as instrument, date of observation, field of view, etc. 
The database also provides the ability to associate custom meta-data to 
each database entry such as keywords, comments, and favouring.

@@

Using the database package, one uses the \texttt{Database} class to connect
with a database and perform operations on it. Listing~\ref{code:database} shows how to connect
to an in-memory database and download data from the \textsc{VSO}. These entries are
automatically added to the database. The function \texttt{len} is used to get the number of
records. The function \texttt{display\_entries} displays an iterable of database entries
in a nicely-formatted \textsc{ASCII} table. The headlines correspond to the
attributes of the respective database entries.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.net import vso
>>> from sunpy.database import Database
>>> database = Database('sqlite:///')
>>> database.download(
...     vso.attrs.Time((2011, 9, 20, 1), (2011, 9, 20, 2)),
...     vso.attrs.Instrument('RHESSI'))
>>> len(database)
6
>>> from sunpy.database.tables import display_entries
>>> print display_entries(
...     database,
...     ['id', 'observation_time_start', 'wavemin', 'wavemax', 'download_time'])
id observation_time_start wavemin        wavemax           download_time      
-- ---------------------- -------        -------           -------------      
1  2011-09-20 01:09:20    0.413280643067 7.29318781883e-05 2014-01-18 19:21:35

>>>>

Easy access to large quantities of solar data frequently leads to datafiles accumulating
in local storage such as laptops and desktop computers. Keeping data organised and available
is typically a cumbersome task for the average user. The file database is a sub-package of 
SunPy which is meant to solve this problem by providing a unified database to store and manage information about 
local data files. The database package was developed as part of the Google 
Summer Of Code (GSOC) program 2013.

The database sub-package makes use of a database supported by
\href{http://www.sqlalchemy.org}{SQLAlchemy}. This library was chosen
since it supports many SQL dialects. 
If SQLite is chosen, the database is stored as a single file which is
created automatically.A server-based database, on the other hand, could be used 
by a collaboration who work together from different machines on the
same data: a central database server stores all data and the clients connect to
it to read or write data.

The database can store and manage all data that can be read via SunPy's 
\texttt{io} subpackage, while direct integration with the \textsc{VSO} 
subpackage is supported.
It also possible to add entries manually by file or directory. The package also provides
a \texttt{fetch} method which provides a unified data search which includes both local files
and files on the \textsc{VSO} are searched. This reduces the likelihood of downloading the same file 
multiple times. When a file is added to the database the file is scanned and a file hash is produced. 
The current date is associated with the entry along with metadata summaries such 
as instrument, date of observation, field of view, etc. 
The database also provides the ability to associate custom meta-data to 
each database entry such as keywords, comments, and favouring, as well as 
querying the full metadata (\textit{i.e.} FITS header) of each entry.

@@

The \texttt{Database} class connects to a database and allows the user to 
perform operations on it. Listing~\ref{code:database} shows how to connect
to an in-memory database and download data from the \textsc{VSO}. These entries are
automatically added to the database. The function \texttt{len} is used to get the number of
records. The function \texttt{display\_entries} displays an iterable of 
database entries in a nicely-formatted \textsc{ASCII} table. The headlines 
correspond to the attributes of the respective database entries.

\begin{listing}[H]
\begin{minted}[bgcolor=bg]{pycon}
>>> from sunpy.net import vso
>>> from sunpy.database import Database
>>> database = Database('sqlite:///')
>>> database.download(
...     vso.attrs.Time((2011, 9, 20, 1), (2011, 9, 20, 2)),
...     vso.attrs.Instrument('RHESSI'))
>>> len(database)
6
>>> from sunpy.database.tables import display_entries
>>> print display_entries(
...     database,
...     ['id', 'observation_time_start', 'wavemin', 'wavemax'])
id observation_time_start wavemin        wavemax    
-- ---------------------- -------        -------    
1  2011-09-20 01:09:20    0.413280643067 7.29318781883e-05
